---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mstate2

<!-- badges: start -->
[![R-CMD-check](https://github.com/jcarmezim/mstate2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jcarmezim/mstate2/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/jcarmezim/mstate2/graph/badge.svg)](https://app.codecov.io/gh/jcarmezim/mstate2)
<!-- badges: end -->

mstate2 extends the classical first-order multistate framework (as in the
**mstate** package) with a *second-order*, short-memory layer: the future
evolution may depend not only on the current state but also on the state
occupied at the immediately preceding time. It provides nonparametric estimators
of the 1-step second-order transition probabilities, their variances and
confidence intervals, an extended Chapman–Kolmogorov relation for n-step
prediction, and evolution intervals for assessing whether — and for how long —
the preceding state matters. The methods are those of Najera-Zuloaga, Besalú and
Gómez Melis (2025), building on Besalú and Gómez Melis (2024).

## Installation

You can install the development version of mstate2 from [GitHub](https://github.com/) with:

```r
# install.packages("pak")
pak::pak("jcarmezim/mstate2")
```

## Example

A second-order model where, from state `B`, the next move depends on whether the
patient arrived from `A` or from `B`. We simulate from a known model, estimate
the transition probabilities, and assess how long the preceding state matters:

```{r example, fig.height = 4, fig.width = 7}
library(mstate2)

## a small second-order model: A -> B -> {B, C}, C absorbing
st   <- c("A", "B", "C")
tens <- array(0, c(3, 3, 3), dimnames = list(st, st, st))
tens["B", "B", "A"] <- 0.6; tens["B", "C", "A"] <- 0.4   # from (A, B)
tens["B", "B", "B"] <- 0.3; tens["B", "C", "B"] <- 0.7   # from (B, B)
tens["C", "C", ]    <- 1                                  # C absorbing
first <- matrix(0, 3, 3, dimnames = list(st, st)); first["A", "B"] <- 1

set.seed(1)
panel <- mstate2::simulate2(5000, tens, first, init = c(A = 1, B = 0, C = 0))
```

Estimate the 1-step second-order transition probabilities with the relative
probability estimator (the recommended default):

```{r estimate}
d   <- prep2(panel, id = "id", time = "time", state = "state")
fit <- P2est(d, estimator = "RPE")
head(fit$estimate)
```

Predict several steps ahead and compare the trajectories for the two preceding
states. `summary()` reports the duration of the second-order effect, and the
plot draws the evolution intervals, marking the first step at which they overlap:

```{r compare, fig.height = 4, fig.width = 7}
cmp <- compare2(fit, h = c("A", "B"), j = "B", l = "B", nsteps = 9)
summary(cmp)

plot(cmp, type = "interval")
```

Non-overlapping evolution intervals imply a statistically significant effect of
the preceding state, so the overlap step quantifies how long the second-order
memory persists.

## Why `README.Rmd`?

Using `README.Rmd` rather than a hand-written `README.md` means the examples
above are executed when the file is knit, so the README cannot drift out of sync
with the code. Re-render it with:

``` r
devtools::build_readme()
```

Remember to commit and push the figure files generated under `man/figures/` so
they display on GitHub.

## References

- J. Najera-Zuloaga, M. Besalú, G. Gómez Melis. *Second-Order Markov Multistate
  Models: Nonparametric Estimation and Inference* (2025).
- M. Besalú, G. Gómez Melis. *Second order Markov multistate models*. SORT (2024).
