---
title: "Second-order Markov multistate models with mstate2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Second-order Markov multistate models with mstate2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(mstate2)
```

## Why second order?

A first-order Markov multistate model assumes the future depends only on the
*current* state. A **second-order** model relaxes this: the next state may also
depend on the *previous* state, capturing short-term memory. `mstate2` estimates
the 1-step second-order transition probabilities

$$P_{hj\ell} = P(X_s = \ell \mid X_{s-1} = j,\; X_{s-2} = h),$$

provides confidence intervals, predicts any number of steps ahead via an
extended Chapman–Kolmogorov relation, constructs *evolution intervals* to test
whether — and for how long — the preceding state matters, tests the
first-order assumption formally, and extends the framework with covariates.
This vignette walks through the whole workflow end to end.

## A worked example

We simulate a three-state process. From state `B` the one-step-ahead behaviour
depends on whether the patient arrived from `A` or from `B`:

```{r}
st   <- c("A", "B", "C")
tens <- array(0, c(3, 3, 3), dimnames = list(st, st, st))
tens["B", "B", "A"] <- 0.6; tens["B", "C", "A"] <- 0.4   # from (A,B): stay 0.6
tens["B", "B", "B"] <- 0.3; tens["B", "C", "B"] <- 0.7   # from (B,B): stay 0.3
tens["C", "C", ]    <- 1                                  # C absorbing
first <- matrix(0, 3, 3, dimnames = list(st, st)); first["A", "B"] <- 1

set.seed(1)
panel <- simulate2(5000, tens, first, init = c(A = 1, B = 0, C = 0))
head(panel)
```

## Estimation

`prep2()` builds the second-order counting processes; `P2est()` estimates the
probabilities with the relative probability estimator (RPE, the recommended
default) or the conditional probability estimator (CPE).

```{r}
d   <- prep2(panel, id = "id", time = "time", state = "state")
fit <- P2est(d, estimator = "RPE")
fit$estimate
```

The estimates recover the truth: `A -> B -> B` near 0.6, `B -> B -> B` near 0.3.

## Prediction

`ckequations()` propagates the 1-step estimates to any horizon:

```{r}
ckequations(fit, h = "A", j = "B", l = "B", nsteps = 6)
```

## Evolution intervals

`compare2()` contrasts the trajectories for different preceding states and
attaches evolution intervals; `summary()` reports the duration of the effect.

```{r}
cmp <- compare2(fit, h = c("A", "B"), j = "B", l = "B", nsteps = 9)
summary(cmp)
```

```{r, fig.alt = "Evolution intervals for the two preceding states"}
plot(cmp, type = "interval")
```

Non-overlapping evolution intervals imply a statistically significant effect of
the preceding state. Here the two trajectories are clearly separated in the
early steps, quantifying how long the second-order memory persists.

## Testing the first-order Markov assumption

`compare2()`'s interval overlap is a visual, pairwise check. `markov_test()`
turns it into a formal test, per (current, target) transition, of whether the
preceding state matters at all — an inverse-variance-weighted heterogeneity
test built directly on the RPE/CPE asymptotics already used above (no extra
model to fit):

```{r}
markov_test(fit)
```

Both transitions out of `B` are highly significant: as simulated, the
preceding state clearly changes the 1-step probabilities.

## Adding covariates

The estimates above are non-parametric and covariate-free: they cannot tell
you whether the risk differs for, say, an older patient. `prep2(covariates =
)` carries baseline covariates through to an individual-level table
(`$triples`), and `P2reg()` fits a discrete-time cause-specific hazard
regression on it — the covariate extension of the second-order framework.

We attach a synthetic, subject-level `age` covariate to the same panel (it
plays no role in how the process was simulated, so this doubles as a sanity
check: its estimated effect should be negligible):

```{r}
ids       <- unique(panel$id)
set.seed(2)
age_by_id <- stats::setNames(sample(40:80, length(ids), replace = TRUE), ids)
panel$age <- age_by_id[as.character(panel$id)]

d2  <- prep2(panel, covariates = "age", states = st)
reg <- P2reg(d2, h = "A", j = "B", formula = ~age)
summary(reg)
```

Under the default complementary log-log link, `exp(coef)` is a hazard ratio —
the discrete-time analogue of a Cox hazard ratio (Prentice & Gloeckler, 1978).
`predict()` gives the covariate-adjusted 1-step probabilities for any profile:

```{r}
predict(reg, newdata = data.frame(age = c(40, 60, 80)))
```

As an internal check, an intercept-only model (`formula = ~1`, no covariates)
must reduce to the same sample proportion of the risk set as the RPE estimate
above, whatever link function is used — and it does, up to the GLM's
convergence tolerance:

```{r}
reg0 <- P2reg(d2, h = "A", j = "B", formula = ~1)
predict(reg0, newdata = data.frame(age = NA))
fit$estimate[fit$estimate$h == "A" & fit$estimate$j == "B", c("l", "p")]
```

## Comparing against a first-order model

A second-order model needs $M$ matrices of dimension $M \times M$ where a
first-order model needs only one. Before paying that cost, `compare_order()`
shows how much (and for how long) knowing the preceding state actually
changes the forecast, relative to a first-order baseline; `divergence()`
reports how many steps that difference survives.

`compare_order()` expects an `mstate::probtrans` object, fit with `predt = 1`
so it starts at the same "time 1" as `ckequations()`. On a real, progressive
multistate dataset (no self-transitions), that looks like:

```{r, eval = FALSE}
library(mstate)
tmat  <- as_tmat(d2)                                    # interoperate with mstate
msdat <- msprep(time = ..., status = ..., data = ..., trans = tmat, keep = "age")
cx    <- coxph(Surv(Tstart, Tstop, status) ~ strata(trans), data = msdat, method = "breslow")
msf   <- msfit(cx, trans = tmat)
pt    <- probtrans(msf, predt = 1)                       # landmark at time 1
compare_order(fit, pt, h = c("A", "B"), j = "B", l = "B", nsteps = 6)
```

Our toy process has a self-transition (`B -> B`), which does not fit
`msprep()`'s usual "each state entered once" progressive-model assumption, so
for this self-contained vignette we instead build the first-order baseline by
*pooling the RPE estimate over the preceding state* — i.e. deliberately
discarding the history mstate2 otherwise keeps — and propagate it with
`ckequations()` on an `h`-independent tensor, wrapped into the structure
`probtrans()` returns. With real data, prefer the `mstate` pipeline above.

```{r, fig.alt = "Second-order curves against a pooled first-order baseline"}
est   <- fit$estimate
denom <- sum(unique(est[est$j == "B", c("h", "at.risk")])$at.risk)
p_BB  <- sum(est$n.trans[est$j == "B" & est$l == "B"]) / denom
p_BC  <- sum(est$n.trans[est$j == "B" & est$l == "C"]) / denom
se_BB <- sqrt(p_BB * (1 - p_BB) / denom)
se_BC <- sqrt(p_BC * (1 - p_BC) / denom)

P1 <- matrix(0, 3, 3, dimnames = list(st, st))
P1["A", "A"] <- 1                                        # never revisited; unused placeholder
P1["B", "B"] <- p_BB; P1["B", "C"] <- p_BC
P1["C", "C"] <- 1
P1_tensor <- array(rep(as.vector(P1), 3), c(3, 3, 3), dimnames = list(st, st, st))

nsteps <- 6
curve  <- ckequations(P1_tensor, h = "A", j = "B", l = NULL, nsteps = nsteps)  # h-independent

pt_B <- data.frame(time = c(1, 1 + seq_len(nsteps)))
mat  <- rbind(c(A = 0, B = 1, C = 0), curve[, st])        # row 1: trivial starting distribution
pt_B$pstate1 <- mat[, "A"]; pt_B$se1 <- 0
pt_B$pstate2 <- mat[, "B"]; pt_B$se2 <- se_BB
pt_B$pstate3 <- mat[, "C"]; pt_B$se3 <- se_BC
pt <- vector("list", 3); pt[[2]] <- pt_B; class(pt) <- "probtrans"

cmp2 <- compare_order(fit, pt, h = c("A", "B"), j = "B", l = "B", nsteps = nsteps)
plot(cmp2, type = "interval")
divergence(cmp2)
```

Because this pooled baseline mixes together the only two histories that feed
into it here, it sits between the two second-order curves rather than
representing an independent, unbiased first-order fit — a limitation of the
shortcut, not of `compare_order()` itself.

## Real data and interoperability

For sojourn-time datasets (such as the DIVINE COVID-19 cohort), use
`sojourn_to_panel()` to obtain panel data, then proceed as above. `as_tmat()`
returns an `mstate::transMat` transition matrix so the model can be read
alongside a first-order **mstate** analysis, and `from_msdata(keep = )` /
`prep2(covariates = )` carry baseline covariates through from an `mstate`
`msdata` object. See the reproduction scripts in the package `tests/`
directory for an end-to-end analysis of the DIVINE data.
