---
title: "Second-order Markov multistate models with mstate2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Second-order Markov multistate models with mstate2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(mstate2)
```

## Why second order?

A first-order Markov multistate model assumes the future depends only on the
*current* state. A **second-order** model relaxes this: the next state may also
depend on the *previous* state, capturing short-term memory. `mstate2` estimates
the 1-step second-order transition probabilities

$$P_{hj\ell} = P(X_s = \ell \mid X_{s-1} = j,\; X_{s-2} = h),$$

provides confidence intervals, predicts any number of steps ahead via an
extended Chapman–Kolmogorov relation, and constructs *evolution intervals* to
test whether — and for how long — the preceding state matters.

## A worked example

We simulate a three-state process. From state `B` the one-step-ahead behaviour
depends on whether the patient arrived from `A` or from `B`:

```{r}
st   <- c("A", "B", "C")
tens <- array(0, c(3, 3, 3), dimnames = list(st, st, st))
tens["B", "B", "A"] <- 0.6; tens["B", "C", "A"] <- 0.4   # from (A,B): stay 0.6
tens["B", "B", "B"] <- 0.3; tens["B", "C", "B"] <- 0.7   # from (B,B): stay 0.3
tens["C", "C", ]    <- 1                                  # C absorbing
first <- matrix(0, 3, 3, dimnames = list(st, st)); first["A", "B"] <- 1

set.seed(1)
panel <- simulate2(5000, tens, first, init = c(A = 1, B = 0, C = 0))
head(panel)
```

## Estimation

`prep2()` builds the second-order counting processes; `P2est()` estimates the
probabilities with the relative probability estimator (RPE, the recommended
default) or the conditional probability estimator (CPE).

```{r}
d   <- prep2(panel, id = "id", time = "time", state = "state")
fit <- P2est(d, estimator = "RPE")
fit$estimate
```

The estimates recover the truth: `A -> B -> B` near 0.6, `B -> B -> B` near 0.3.

## Prediction

`ckequations()` propagates the 1-step estimates to any horizon:

```{r}
ckequations(fit, h = "A", j = "B", l = "B", nsteps = 6)
```

## Evolution intervals

`compare2()` contrasts the trajectories for different preceding states and
attaches evolution intervals; `summary()` reports the duration of the effect.

```{r}
cmp <- compare2(fit, h = c("A", "B"), j = "B", l = "B", nsteps = 9)
summary(cmp)
```

```{r, fig.alt = "Evolution intervals for the two preceding states"}
plot(cmp, type = "interval")
```

Non-overlapping evolution intervals imply a statistically significant effect of
the preceding state. Here the two trajectories are clearly separated in the
early steps, quantifying how long the second-order memory persists.

## Real data and interoperability

For sojourn-time datasets (such as the DIVINE COVID-19 cohort), use
`sojourn_to_panel()` to obtain panel data, then proceed as above. `as_tmat()`
returns an `mstate::transMat` transition matrix so the model can be read
alongside a first-order **mstate** analysis. See the reproduction scripts in the
package `tests/` directory for an end-to-end analysis of the DIVINE data.
